A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the span over which ancestral haplotypes extend
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the Robinson-Foulds distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct, % NSP-22NOV: "vast majority" is correct in simulated (but not inferred) ARGs, right?
% HF-24NOV : It should be both, as the IE line in Figure 6D shows that inferred trees match the TPR of the ground truth by at least 60% for inferred trees.
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

>>>>

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the span over which ancestral haplotypes extend
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the Robinson-Foulds distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct if the trees are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.
